ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}

### 2021-01-03 manuscript/main.tex
manuscript: update to latest text, scripts, figures and results data

\title{ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
\author[1,2,3,4]{David Lähnemann}
\author[4,5]{Johannes Köster}
\author[3]{Ute Fischer}
\author[3]{Arndt Borkhardt}
\author[1,2,+,*]{Alice C. McHardy}
\author[5,6,+,*]{Alexander Schönhuth}
\affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
\affil[2]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[3]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[4]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
\affil[5]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
\affil[6]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
\affil[+]{shared last authors}
\affil[*]{To whom correspondence should be addressed:
 \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de},
 \href{as@cwi.nl}{as@cwi.nl}.}

@@

\section*{\Large Abstract}
{\sffamily
Obtaining accurate mutational profiles from single cell DNA is essential for the analysis of genomic cell-to-cell heterogeneity at the finest level of resolution.
However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which when applied to single cells generate significant numbers of false positives and false negatives.
Only dedicated models accounting for amplification bias and errors will be able to provide more accurate calls.

We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified single cell DNA sequencing data.
It introduces a mechanistically motivated empirical model of amplification bias that improves the quantification of genotyping uncertainty.
To account for amplification errors, it jointly models the single cell sample with a bulk sequencing sample from the same cell population---also enabling a biologically relevant imputation of missing genotypes for the single cell.
Through these innovations, ProSolo achieves substantially higher performance in calling and genotyping single nucleotide variants in single cells in comparison to all state-of-the-art tools. 
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to identify, such as allele dropout.

@@

Originally, genome sequences have been queried for genetic germline variation or for highly abundant somatic variation, for example in cancer.
The advent of high-throughput single cell sequencing has recently turned the spotlight on a type of so far understudied variation: 
the often less abundant somatic or post-zygotic variation that constantly accumulates with every mitotic cell division throughout the lifetime of an organism, turning every individual into a complicated genomic mosaic\cite{forsberg_mosaicism_2017}.
Estimates for somatic single nucleotide variants (SNVs) range from around $0.6 \cdot 10^{-9}$  up to $60 \cdot 10^{-9}$  mutations per genome position per cell division\cite{lynch_rate_2010,lynch_evolution_2010,tomasetti_half_2013,li_somatic_2013}, with a recent estimate\cite{milholland_differences_2017} based on single-cell sequencing at $2.66 \cdot 10^{-9}$.
With the size of the human (reference) genome at approximately $3.2 \cdot 10^9$ base pairs, these numbers indicate that even during healthy development, most cells harbour cell-specific point mutations.
This enables retrospective monitoring of lineages involved in normal organism development, merely by sampling some cells\cite{kester_single-cell_2018}, without having to interfere with its general development, or kill the individual.
In other words, this establishes a universally applicable methodology for \textit{in vivo} lineage tracing.

@@

Thus, variant callers for whole-genome-amplified single-cell data need to account for both amplification errors and allelic bias, in addition to accounting for the site-specific variation.
However, state of the art single cell SNV callers routinely assume fixed global rates when modeling uneven allelic coverage (up to dropout) and amplification errors. 
For example, to reflect the amplification error rate, both MonoVar\cite{zafar_monovar:_2016} and SCcaller\cite{dong_accurate_2017} work with global false positive error rates for calling the presence of an alternative allele at a particular site.
This assumes that $\Phi29$ polymerase is agnostic to local template sequence context, although it is not\cite{de_paz_high-resolution_2018}.
Similarly, for modelling allele dropout, MonoVar\cite{zafar_monovar:_2016} and SCIPhI\cite{singer_single-cell_2018} assume that one rate applies globally (and across all cells).
This neglects that allele dropout, as the extreme case of uneven allele coverage, varies greatly along the genome and in particular also between cells, because their DNA is amplified separately.
Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
Improving on that point, tools exist that account for local variation in allelic amplification bias. SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019} both estimate the minor allele frequency from nearby germline heterozygous sites. 
However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
Thus, to the best of our knowledge, there is no statistical model that allows for local variation of bias and errors due to amplification, and for statistically sound false discovery control when calling and genotyping SNVs in single cells.

We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
Importantly, our model further enables a computationally efficient implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.

@@

\begin{figure}[!tpb]%figure1
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_hom-ref-model_cov-20.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_het-model_cov-20.pdf}
    \end{minipage}
  \end{minipage}
  \vspace{2ex}
  \hspace{1ex}
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/prosolo_alt-calling_overview.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.49\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Event_space.pdf}
    \end{minipage}
  \end{minipage}
  \caption{
   \textbf{A,B} Exemplary alternative allele read count distributions for sites covered by 20 reads, as derived by Lodato et al.\cite{lodato_somatic_2015}
   Homozygous reference sites in \textbf{(A)} are assumed to follow a beta-binomial distribution;
   sites heterozygous for the alternative allele in \textbf{(B)} are assumed to follow the linear combination of two symmetrical beta-binomial distributions (dotted and dashed lines).
   \textbf{(C)} Toy example of calling the same genomic site in two single cells from the same population that differ in their true underlying allele frequencies for alternative allele C (blue, $\theta_s = 0$ vs. $\theta_s = 0.5$).
   Alternative nucleotide T (orange) is an amplification error.
   Empirical distributions in A and B account for the  amplification bias, and likelihoods for the alternative allele candidates from the bulk reduce the likelihoods of amplification errors, thereby correctly identifying both the error and the original true mutation.
   This is formalised with the model in D.
   \textbf{(D)} Definition of single cell events based on true underlying alternative allele frequencies in the single cell ($\theta_s$) and the bulk ($\theta_b$) (assuming that the bulk sample has a deep coverage that captures somatic variants).
   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).\small\newline
   ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
  \label{fig:prosolo_alt-calling}
\end{figure}

@@

the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}C).
We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele.

@@

For our model, we derive likelihoods for all possible alternative allele frequencies in the background bulk sample.
Given a set of $n$ reads from the bulk ($b$) read data $\boldsymbol{Z}^b  = \{ \boldsymbol{Z}_1^b, \boldsymbol{Z}_2^b, \dots, \boldsymbol{Z}_n^b \}$, and discrete possible allele frequencies $\frac{m}{n}$ ($m \in 0, 1, \dots, n$), we compute the probability of the data given a particular allele frequency as the product of the probabilities of all the reads:

@@

\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihoods works for calling genotypes in the single cell.
However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:

\begin{equation}
  \label{eq:event-space}
  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\theta_s} \times [0, 1]_{\theta_b}.
\end{equation}

Within this space, we define single cell events as mutually exclusive subspaces, for example an error-free homozygous alternative site is defined by allele frequencies $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\theta_s} \times (\frac{1}{2}, 1]_{\theta_b}$  or the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}D):

@@

We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}. We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 5.4.0}) to implement the benchmarking workflows.
For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769116}).

@@

\begin{figure}[!tpb]
  \centering
  \begin{minipage}{.26\linewidth}
    \includegraphics[height=32ex]{figs/Dong2017_data_overview.pdf} \\
  \end{minipage}
  \hspace{4ex}
  \begin{minipage}{.4\linewidth}
    \includegraphics[height=32ex]{figs/Laehnemann2017_data_overview.pdf} \\
  \end{minipage}
  \hspace{4ex}
  \begin{minipage}{.22\linewidth}
    \includegraphics[height=32ex]{figs/data_overview_color_legend.pdf} \\
  \end{minipage}
  \caption{
   Benchmark Datasets.
   For single cells, DNA was multiple displacement amplified (MDA).
   \textbf{A} Whole genome sequencing (WGS) dataset, generated from a clonal population started with an individual cell\cite{dong_accurate_2017}, and expanded further downstream to generate different bulk samples (C1, C2, C3, IL-1C).
   \textbf{B} Newly generated whole exome sequencing (WXS) dataset of blood cells.
   Ground truth genotypes of patient CCS1 were determined from sequencing data of family members (boxes male; circles female).
   Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
   }
  \label{fig:datasets}
\end{figure}

We benchmarked ProSolo on two experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\

@@

\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}B, dataset available as \mbox{EGAD00001005929} in EGA study \mbox{EGAS00001004123}).
% refs for MACS and FACS? Ute said, this is usually not required.
Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
Using a panel of 16 primer pairs covering different genes across chromosomes for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
% ref for q-rt-PCR? Ute said, this is usually not required.
For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.

@@

\subsubsection{ProSolo achieves highest alternative allele calling accuracy}

@@

\begin{figure}[!tpb]
 \begin{minipage}{.40\linewidth}
  \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}
 \begin{minipage}{.55\linewidth}
  \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
 \end{minipage}

>>>>

\title{ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
\author[1,2,3,4]{David Lähnemann}
\author[4,5]{Johannes Köster}
\author[3]{Ute Fischer}
\author[3]{Arndt Borkhardt}
\author[1,2,+,*]{Alice C. McHardy}
\author[5,6,+,*]{Alexander Schönhuth}
\affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
\affil[2]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[3]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[4]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
\affil[5]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
\affil[6]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
\affil[+]{shared last authors}
\affil[*]{To whom correspondence should be addressed:
 \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de},
 \href{as@cwi.nl}{as@cwi.nl}.}

@@

\section*{\Large Abstract}
{\sffamily
Obtaining accurate mutational profiles from single cell DNA is essential for the analysis of genomic cell-to-cell heterogeneity at the finest level of resolution.
However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which---when applied to single cells---generate significant numbers of false positives and false negatives.
Only dedicated models accounting for amplification bias and errors will be able to provide more accurate calls.

We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
ProSolo models single cells jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered polymorphic sites.
As a consequence, ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to identify, such as allele dropout.

@@

Originally, genome sequences have been queried for genetic germline variation or for highly abundant somatic variation, for example in cancer.
The advent of high-throughput single cell sequencing has recently turned the spotlight on a type of so far understudied variation: 
the often less abundant somatic or post-zygotic variation that constantly accumulates with every mitotic cell division throughout the lifetime of an organism, turning every individual into a complicated genomic mosaic\cite{forsberg_mosaicism_2017}.
Estimates for somatic single nucleotide variants (SNVs) range from around $0.6 \cdot 10^{-9}$  up to $60 \cdot 10^{-9}$  mutations per genome position per cell division\cite{lynch_rate_2010,lynch_evolution_2010,tomasetti_half_2013,li_somatic_2013}, with a recent estimate\cite{milholland_differences_2017} based on single-cell sequencing at $2.66 \cdot 10^{-9}$.
With the size of the human (reference) genome at approximately $3.2 \cdot 10^9$ base pairs, these numbers indicate that even during healthy development, most cells harbour cell-specific point mutations.
This enables retrospective monitoring of lineages involved in normal organism development, merely by sampling some cells\cite{kester_single-cell_2018}, without having to interfere with its general development or having to kill the individual.
In other words, this establishes a universally applicable methodology for \textit{in vivo} lineage tracing.

@@

Thus, variant callers for whole-genome-amplified single-cell data need to account for both amplification errors and allelic bias, in addition to accounting for the site-specific variation.
However, state of the art single cell SNV callers routinely assume fixed global rates when modeling uneven allelic coverage (up to dropout) and amplification errors. 
For example, to reflect the amplification error rate, both MonoVar\cite{zafar_monovar:_2016} and SCcaller\cite{dong_accurate_2017} work with global false positive error rates for calling the presence of an alternative allele at a particular site.
This assumes that $\Phi29$ polymerase is agnostic to local template sequence context, although it is not\cite{de_paz_high-resolution_2018}.
Similarly, for modelling allele dropout, MonoVar\cite{zafar_monovar:_2016} and SCIPhI\cite{singer_single-cell_2018} assume that one rate applies globally (and across all cells).
This neglects that allele dropout, as the extreme case of uneven allele coverage, varies greatly along the genome and in particular also between cells, because their DNA is amplified separately.
Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
Two tools that more variably model alleleic amplification bias are SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019}.
Both estimate the minor allele frequency from nearby germline heterozygous sites. 
However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
Thus, to the best of our knowledge, there is no statistical model that allows for local variation of bias and errors due to amplification, and for statistically sound false discovery control when calling and genotyping SNVs in single cells.

We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
Importantly, our model enables a scalable---because computationally efficient---implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.

@@

\begin{figure}[!tpb]%figure1
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_hom-ref-model_cov-20.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_het-model_cov-20.pdf}
    \end{minipage}
  \end{minipage}
  \vspace{2ex}
  \hspace{1ex}
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/prosolo_alt-calling_overview.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.49\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Event_space.pdf}
    \end{minipage}
  \end{minipage}
  \caption{
   \textbf{A,B} Exemplary alternative allele read count distributions for sites covered by 20 reads, as derived by Lodato et al.\cite{lodato_somatic_2015}
   Homozygous reference sites in \textbf{(A)} are assumed to follow a beta-binomial distribution;
   sites heterozygous for the alternative allele in \textbf{(B)} are assumed to follow the linear combination of two symmetrical beta-binomial distributions (dotted and dashed lines).
   \textbf{(C)} Toy example of calling the same genomic site in two single cells from the same population that differ in their true underlying allele frequencies for alternative allele C (blue, $\theta_s = 0$ vs. $\theta_s = 0.5$).
   Alternative nucleotide T (orange) is an amplification error.
   Empirical distributions in A and B account for the  amplification bias, and likelihoods for the alternative allele candidates from the bulk reduce the likelihoods of amplification errors, thereby correctly identifying both the error and the original true mutation.
   This is formalised with the model in D.
   \textbf{(D)} Definition of single cell events based on ProSolo's likelihood density estimates for the spectrum of true underlying alternative allele frequencies in the single cell ($\tilde{\theta}_s$) and the bulk ($\tilde{\theta}_b$).
   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
   The model further assumes that the bulk sample has sufficient coverage to capture somatic variants.\small\newline
   ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
  \label{fig:prosolo_alt-calling}
\end{figure}

@@

the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}C).
We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele, to calculate likelihoods for each of these possible true allele frequencies.

@@

For our model, we derive likelihood density estimates for all possible alternative allele frequencies in the background bulk sample.
Given a set of $n$ reads from the bulk ($b$) read data $\boldsymbol{Z}^b  = \{ \boldsymbol{Z}_1^b, \boldsymbol{Z}_2^b, \dots, \boldsymbol{Z}_n^b \}$, and discrete possible allele frequencies $\frac{m}{n}$ ($m \in 0, 1, \dots, n$), we compute the probability of the data given a particular allele frequency as the product of the probabilities of all the reads:

@@

\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across allele frequency $\tilde{\theta}_s,\tilde{theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:

\begin{equation}
  \label{eq:event-space}
  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\tilde{\theta}_s} \times [0, 1]_{\tilde{\theta}_b}.
\end{equation}

For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
For example, an error-free homozygous alternative site is defined by allele frequency likelihood estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
And the dropout of an alternative allele is defined across allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.

We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}D):

@@

We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}. We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 5.4.0}) to implement the benchmarking workflows.
For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3824152}).

@@

\begin{figure}[!tpb]
  \begin{minipage}{.3\linewidth}
    \includegraphics[height=38ex]{figs/Dong2017_data_overview.pdf} \\
  \end{minipage}
  \hspace{6ex}
  \begin{minipage}{.5\linewidth}
    \includegraphics[height=38ex]{figs/Laehnemann2017_data_overview.pdf} \\
  \end{minipage}\\
  \begin{minipage}{.5\linewidth}
    \includegraphics[height=38ex]{figs/Wang2014_data_overview.pdf} \\
  \end{minipage}
  \hspace{6ex}
  \begin{minipage}{.3\linewidth}
    \includegraphics[height=38ex]{figs/data_overview_color_legend.pdf} \\
  \end{minipage}
  \caption{
   Benchmark Datasets.
   For single cells, DNA was multiple displacement amplified (MDA).
   \textbf{A} Whole genome sequencing (WGS) dataset, generated from a clonal population started with an individual cell\cite{dong_accurate_2017}, and expanded further downstream to generate different bulk samples (C1, C2, C3, IL-1C).
   \textbf{B} Newly generated whole exome sequencing (WXS) dataset of blood cells.
   Ground truth genotypes of patient CCS1 were determined from sequencing data of family members (boxes male; circles female).
   Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
   \textbf{C} Whole exome sequencing (WXS) dataset from a triple negative breast cancer (TNBC) patient\cite{wang_clonal_2014}.
   For tumor cells (a1 to a8 and h1 to h8), the cell population (pop) sequencing of the normal sample was taken as the ground truth, including known clonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
   For normal cells (n1 to n16), the cell population (pop) sequencing of the tumor sample was taken as the ground truth, after removing known clonal and subclonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
   }
  \label{fig:datasets}
\end{figure}

We benchmarked ProSolo on three experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\

@@

\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}B, dataset available as \mbox{EGAD00001005929} in EGA study \mbox{EGAS00001004123}).
Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
Using a panel of 16 primer pairs covering different genes across chromosomes for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.

@@

\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC)\cite{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available (Figure~\ref{fig:datasets}C, dataset available via the sequence read archive, project id \mbox{PRJNA168068} or accession \mbox{SRA053195}).
For each, 16 single cell nuclei and a bulk sample were whole exome sequenced.
Further, a number of candidate sites for clonal and subclonal tumor mutations was confirmed by targeted duplex deep sequencing in a separate experiment\cite{wang_clonal_2014}.
We analyzed the tumor and normal cells separately, ensuring that normal cells were called with the normal bulk background sample and tumor cells with the tumor bulk background sample.
As the ground truth for the tumor cells, we then used the normal bulk sample augmented with the clonal tumor mutations confirmed by targeted duplex sequencing.
As the ground truth for the normal cells, we used the tumor bulk sample, removing the confirmed clonal and subclonal tumor mutations. Note that this experimental setup accounts for utmost fairness for all competitors. 
Variant calling for the ground truths was performed as for cell line data above.

\subsubsection{ProSolo achieves consistently high alternative allele calling accuracy}

@@

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
 \end{minipage}\newline
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=40ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}

 \caption{
  Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
  Note that all panels are strong zoom-ins, focusing on (different) areas of interest.
  Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B. \newline \footnotesize
  Threshold parameters:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
\label{fig:alt-calling_prec-rec}
\end{figure}
